#Economic and Health Impact of Extreme Weather On U.S., 1950-2011
##Synopsis
Each year, natural disasters cause millions of dollars in damage to
American property, and cause injuries and deaths. Using the data from the
U.S. National Oceanic and Atmospheric Administration's storm database, we have
conducted an exploratory analysis to determine the most harmful of these
events.

We find that tornadoes are the most destructive to human life, causing the
greatest number of deaths and injuries. Drought causes the most damage to
American crops, and flooding causes the greatest amout of damage to
American property.

##Data Processing
Code is included to download the raw data, if the file is not in the current
working directory. Then, the file is read directly by read.csv. We will also
be using the dplyr and ggplot2 packages; these are loaded.

```{r}
if (!file.exists("repdata%2Fdata%2FStormData.csv.bz2")) {
  download.file(
    "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
    destfile = "repdata%2Fdata%2FStormData.csv.bz2")
}
storm.data <- read.csv("repdata%2Fdata%2FStormData.csv.bz2", stringsAsFactors = FALSE)
library(dplyr)
library(ggplot2)
```

For processing the data, we will use the dplyr package and a helper function
called "cost.multiplier," which will be used to map the symbolic multipliers
in the data set (e.g. "K" for one thousand) to the corresponding numbers, so
that it is possible to calculate the economic damage.

```{r}
cost.multiplier <- function(x) {
  if (x == "") {
    0
  }
  else{
    switch(x,
           "b" =,
           "B" = 1000000000,
           "h" =,
           "H" = 100,
           "K" =,
           "k" = 1000,
           "m" =,
           "M" = 1000000,
           "?" =,
           "-" = 0,
           "+" = 1,
           10) 
  }
}
storm.data$PROPDMGMULT <- sapply(storm.data$PROPDMGEXP, cost.multiplier)
storm.data$CROPDMGMULT <- sapply(storm.data$CROPDMGEXP, cost.multiplier)
```

A data frame tbl is created from the data frame, and the columns we will need
are extracted. We will only requier the data on fatalities, injuries, property
and crop damage (and the corresponding multipliers). The actual economic damage
is calculated by multiplying the value in PROPDMG (likewise, CROPDMG) with the
associated value in PROPDMGMULT (CROPDMGMULT).

```{r}
storm.data <- tbl_df(storm.data)
storm.data <- storm.data %>% 
  select(EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGMULT,
         CROPDMG, CROPDMGMULT) %>%
  mutate(PROPCOST = PROPDMG * PROPDMGMULT,
         CROPCOST = CROPDMG * CROPDMGMULT)
head(storm.data)
```

Next, the data are grouped by EVTYPE -- the kind of event. Sums are run to find
the total number of deaths, of injuries, and the economic cost to property and
to crops for each category of event.

```{r}
grouped.data <- group_by(storm.data, EVTYPE)
property.damage <- grouped.data %>%
  summarize(EXPENSE = sum(PROPCOST)) %>%
  arrange(desc(EXPENSE))
crop.damage <- grouped.data %>%
  summarize(EXPENSE = sum(CROPCOST)) %>%
  arrange(desc(EXPENSE))
fatalities <- grouped.data %>%
  summarize(DEATHS = sum(FATALITIES)) %>%
  arrange(desc(DEATHS))
injuries <- grouped.data %>%
  summarize(INJURED = sum(INJURIES)) %>%
  arrange(desc(INJURED))
```

##Results
###Fatalities

An examination of the fatality data shows us that tornadoes are by far, the
most deadly of the tracked storms. Excessive heat are second, with flash
flood a distant third. The ten most deadly tracked categries are charted,
for purposes of comparison.

```{r}
head(fatalities)
fatality.plot <- ggplot(data = head(fatalities, 10),
                      aes(x = reorder(EVTYPE, DEATHS), y = DEATHS)) +
  geom_col() +
  labs(x="Event Type",
       y="Fatalities",
       title="Fatalities from Extreme Weather") +
  coord_flip()
print(fatality.plot)
```

###Injuries

An examination of our injury data again shows tornadoes as the most harmful
of tracked phenomena, with more than ten times a many injuries as TSTM
(thunderstorm) wind, which is second.

```{r}
head(injuries)
```

###Damage to Property

We find that floods have been most destructive to property -- more than twice
as damaging as hurricanes and typhoons. Again, the top ten are charted for
demonstration.

```{r}
head(property.damage)
property.plot <- ggplot(data = head(property.damage, 10),
                  aes(x = reorder(EVTYPE, EXPENSE), y = EXPENSE)) + 
  geom_col() +
  labs(x="Event Type",
       y="Damage (USD)",
       title="Property Damage from Extreme Weather") +
  coord_flip()
print(property.plot)
```

###Damage to Crops

Though various floods and freezes are individually damaging, the single most
destructive extreme weather event to crops is found to be drought.

```{r}
head(crop.damage)
```