---
title: 'Statistical Inference Course Project Part Two: Data Analysis'
author: "John Higgins"
date: "December 3, 2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Overview
We will explore the effects of vitamin supplementation on guinea pig tooth
growth. Supplementation is provided either by ascorbic acid (VC) or by orange
juice (OJ). Three dosage levels were applied: 0.5 mg/day, 1 mg/day, 2 mg/day.
Subseqeunt tooth length is given in unspecified units.

##Examining The Data
We load in the data and acquaint ourselves with its basic details.

```{r}
library(datasets)
data("ToothGrowth")
str(ToothGrowth)
head(ToothGrowth)
summary(ToothGrowth)
```

Let us look at aggregate summary data to determine how supplement and dose
might affect tooth length.

```{r}
aggregate(len ~ supp + dose, data = ToothGrowth, mean)
```

The data suggest that tooth growth is dose-dependent, and for lower doses,
orange juice is superior to ascorbic acid. This effect decreases as dose
increases. Let us visualize the data to see if it strengthens this hypothesis.

```{r}
library(ggplot2)
ggplot(ToothGrowth,
        aes(x = dose, y = len)) +
  geom_point(aes(col = supp)) + 
  geom_smooth(aes(col = supp),
              method = "lm",
              se = FALSE) +
  facet_wrap(~supp, nrow = 1)
```

Indeed, we see that both regression lines terminate at a similar value for
dose 2 mg/day, but the line for ascorbic acid is steeper. Let us test these
hypotheses.

##Hypothesis Testing
Our tentative hypotheses are these:
For a given supplement, greater dose implies greater tooth growth.
Between supplements, OJ produces greater growth than VC at lower doses, but
this difference vanishes as the dose increases.

To check these hypotheses, we will run T-tests. We will compare a supplement
to itself at increasing dosage, and we will compare one supplement to the other
at identical doses. Our null hypothesis is that the means of tooth length will
be identical across supplement and dosage.

###Dose Dependency
####OJ, 0.5 mg/day vs. 1 mg/day
Alternative hypothesis: the mean tooth length for 1 mg/day is greater than that
for 0.5 mg/day.

```{r}
oj.halfmg <- with(ToothGrowth, len[supp == "OJ" & dose == 0.5])
oj.onemg <- with(ToothGrowth, len[supp == "OJ" & dose == 1.0])
t.test(oj.onemg, oj.halfmg, alternative = "greater")
```

For a p-value of .05, the null hypothesis is rejected.

####OJ, 1 mg/day vs. 2 mg/day
Alternative hypothesis: the mean tooth length for 2 mg/day is greater than that
for 1 mg/day.

```{r}
oj.twomg <- with(ToothGrowth, len[supp == "OJ" & dose == 2.0])
t.test(oj.twomg, oj.onemg, alternative = "greater")
```

For p = .05, the null hypothesis is rejected.

####VC, 0.5 mg/day vs. 1 mg/day
Alternative hypothesis: the mean tooth length for 1 mg/day is greater than that
for 0.5 mg/day.

```{r}
vc.halfmg <- with(ToothGrowth, len[supp == "VC" & dose == 0.5])
vc.onemg <- with(ToothGrowth, len[supp == "VC" & dose == 1.0])
t.test(vc.onemg, vc.halfmg, alternative = "greater")
```

For p = .05, the null hypothesis is rejected.

####VC, 1 mg/day vs. 2 mg/day
Alternative hypothesis: the mean tooth length for 2 mg/day is greater than that
for 1 mg/day.

```{r}
vc.twomg <- with(ToothGrowth, len[supp == "VC" & dose == 2.0])
t.test(vc.twomg, vc.onemg, alternative = "greater")
```

For p = .05, the null hypothesis is rejected.

####Conclusion
Supplement dose correlates strongly with tooth growth. An increase in
supplementation level, regardless of the supplement, corresponds to greater
tooth growth, p < .05.

###Supplement Choice
####OJ vs. VC, 0.5 mg/day
Alternative hypothesis: OJ elicits greater tooth growth than VC.

```{r}
t.test(oj.halfmg, vc.halfmg, alternative = "greater")
```

For p = .05, the null hypothesis is rejected.

####OJ vs. VC, 1 mg/day
Alternative hypothesis: OJ elicits greater tooth growth than VC.

```{r}
t.test(oj.onemg, vc.onemg, alternative = "greater")
```

For p = .05, the null hypothesis is rejected.

####OJ vs. VC, 2 mg/day
Alternative hypothesis: OJ elicits greater tooth growth than VC.

```{r}
t.test(oj.twomg, vc.twomg, alternative = "greater")
```

For p = .05, we fail to reject the null hypothesis.

Let us see if there is any detectable difference in mean between the two
supplements, with a two-sided T-test.

```{r}
t.test(oj.twomg, vc.twomg, alternative = "two.sided")
```

Again, for p = .05 or for any reasonable p-value, we fail to reject the null
hypothesis.

####Conclusion
For the two smaller dosages, OJ elicits greater tooth growth than VC (p < .05).
For a dose of 2 mg/day, there is no difference between supplements.

##Summary
Tooth growth increases as supplement growth increases. For a dose of 0.5 mg/day
or 1 mg/day, orange juice results in greater tooth growth than ascorbic acid,
however at a dose of 2 mg/day the tooth growth between supplements is
identical.