---
title: 'Statistical Inference Course Project Part Two: Data Analysis'
author: "John Higgins"
date: "December 3, 2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Overview
We will explore the effects of vitamin supplementation on guinea pig tooth
growth. Supplementation is provided either by ascorbic acid (VC) or by orange
juice (OJ). Three dosage levels were applied: 0.5 mg/day, 1 mg/day, 2 mg/day.
Subseqeunt tooth length is given in unspecified units.

##Examining The Data
We load in the data and acquaint ourselves with its basic details.

```{r}
library(datasets)
data("ToothGrowth")
```

Let us look at aggregate summary data to determine how supplement and dose
might affect tooth length.

```{r}
aggregate(len ~ supp + dose, data = ToothGrowth, mean)
```

The data suggest that tooth growth is dose-dependent, and for lower doses,
orange juice is superior to ascorbic acid. This effect decreases as dose
increases. Let us visualize the data to see if it strengthens this hypothesis.

```{r, fig.height = 3}
library(ggplot2)
ggplot(ToothGrowth,
        aes(x = dose, y = len)) +
  geom_point(aes(col = supp)) + 
  geom_smooth(aes(col = supp),
              method = "lm",
              se = FALSE) +
  facet_wrap(~supp, nrow = 1)
```

Indeed, we see that both regression lines terminate at a similar value for
dose 2 mg/day, but the line for ascorbic acid is steeper. Let us test these
hypotheses.

##Hypothesis Testing
Null hypotheses: tooth growth does not vary with changes in dose or between
supplements.

###Dose Dependency
####OJ
Alternative hypothesis: greater dose elicits greater tooth growth.

```{r}
oj.halfmg <- with(ToothGrowth, len[supp == "OJ" & dose == 0.5])
oj.onemg <- with(ToothGrowth, len[supp == "OJ" & dose == 1.0])
t.test(oj.onemg, oj.halfmg, alternative = "greater")$p.value
oj.twomg <- with(ToothGrowth, len[supp == "OJ" & dose == 2.0])
t.test(oj.twomg, oj.onemg, alternative = "greater")$p.value
```

For p = .05, the null hypothesis is rejected. 1 mg/day elicits greater growth
than 0.5 mg/day; 2 mg/day elicits greater growth than 1 mg/day.

####VC
Alternative hypothesis: greater dose elicits greater tooth growth.

```{r}
vc.halfmg <- with(ToothGrowth, len[supp == "VC" & dose == 0.5])
vc.onemg <- with(ToothGrowth, len[supp == "VC" & dose == 1.0])
t.test(vc.onemg, vc.halfmg, alternative = "greater")$p.value

vc.twomg <- with(ToothGrowth, len[supp == "VC" & dose == 2.0])
t.test(vc.twomg, vc.onemg, alternative = "greater")$p.value
```

For p = .05, the null hypothesis is rejected. 1 mg/day elicits greater growth
than 0.5 mg/day; 2 mg/day elicits greater growth than 1 mg/day.

####Conclusion
Supplement dose correlates strongly with tooth growth. An increase in
supplementation level, regardless of the supplement, corresponds to greater
tooth growth, p < .05.

###Supplement Choice
####OJ vs. VC, 0.5 mg/day
Alternative hypothesis: OJ elicits greater tooth growth than VC.

```{r}
t.test(oj.halfmg, vc.halfmg, alternative = "greater")$p.value
t.test(oj.onemg, vc.onemg, alternative = "greater")$p.value
t.test(oj.twomg, vc.twomg, alternative = "greater")$p.value
```

For p = .05 and dose 0.5 mg/day or 1 mg/day, we reject the null
hypothesis. For dose 2 mg/day, we fail to reject the null hypothesis.

Let us see if there is any detectable difference in mean between the two
supplements at 2 mg/day, with a two-sided T-test.

```{r}
t.test(oj.twomg, vc.twomg, alternative = "two.sided")$p.value
```

Again, for p = .05 or for any reasonable p-value, we fail to reject the null
hypothesis.

####Conclusion
For the two smaller dosages, OJ elicits greater tooth growth than VC (p < .05).
For a dose of 2 mg/day, there is no difference between supplements.

##Summary
Tooth growth increases as supplement growth increases. For a dose of 0.5 mg/day
or 1 mg/day, orange juice results in greater tooth growth than ascorbic acid,
however at a dose of 2 mg/day the tooth growth between supplements is
identical.

We assume that the data are independent and identically distributed.